---
title: "Philadelphia Property Tax Delinquency"
author: "Joy Payton"
date: "11/30/2016"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
```

## Load Libraries

```{r}
library(dplyr)
```

## Obtain Tax Data 

We can get property tax delinquency data from Philadelphia's Open Data site:

```{r cars}
csv <- "https://data.phila.gov/api/views/4v5p-qi2q/rows.csv?accessType=DOWNLOAD&bom=true"
taxData <- read.csv(csv, stringsAsFactors = FALSE)
```

## Clean Data

Remove columns that don't interest us:

```{r}
taxDataCleaned <- taxData %>% select(-c(Unit,Lien.Number))
```

Remove rows that won't help us because they list no owner:

```{r}
taxDataCleaned <- taxDataCleaned %>% filter(Owner != "")
```

Clean out the $ found in the money fields and convert to numeric

```{r}
taxDataCleaned$Principal <- as.numeric(gsub('\\$','',taxDataCleaned$Principal))
taxDataCleaned$Interest <- as.numeric(gsub('\\$','',taxDataCleaned$Interest))
taxDataCleaned$Penalty <- as.numeric(gsub('\\$','',taxDataCleaned$Penalty))
taxDataCleaned$Other <- as.numeric(gsub('\\$','',taxDataCleaned$Other))
taxDataCleaned$Total <- as.numeric(gsub('\\$','',taxDataCleaned$Total))
```

Get just 2000-2015 tax amount owed.  2016 represents bills that may not truly
be in delinquency yet, and prior to 2000 is too far back for us to be interested.

```{r}
lastFifteenData <- taxDataCleaned %>% filter(Tax.Period < 2016, Tax.Period >= 2000)
```

## Who are our owners?

As a first exploration into our data, we'll want to see who owes the most taxes,
in order to understand a bit more about the delinquency landscape in Philadelphia.
Let's limit our investigation to the last year - 2015.  We'll take only those
properties that have overdue tax and do some summarizing.

```{r}
lastFifteenData <- taxDataCleaned %>% filter(Tax.Period < 2016, Tax.Period >= 2000)


totalOwedByOwner <- lastFifteenData %>% 
                    filter(Tax.Period == 2015, Total >0) %>% 
                    group_by(Owner) %>% 
                    summarise(TotalOwed = sum(Total), 
                              NumOverdueProperties = n())
```

Who are the worst offenders?

```{r}
worstOffenders <- totalOwedByOwner %>% arrange(desc(TotalOwed))
as.data.frame(worstOffenders[1:100,])
```

Let's remove governmental properties.  We want to focus on private owners!

```{r}
worstOffenders <- worstOffenders %>% filter(Owner != "PHILA HOUSING AUTHORITY", 
                                            Owner != "CITY OF PHILADELPHIA",
                                            Owner != "ADMIN OF VET AFFAIRS")
```

Let's set some data up for some fun visualizations.  We'll create a CSV and
download it, then load it up in Joy's Google Drive.  That makes it easy to use
Google Charts to create our "2015 Hall of Shame" for major delinquents.

```{r}
write.csv(worstOffenders[1:100,],"worstOffenders.csv", row.names = FALSE)
```

Where did our data end up?

I stashed it [on Github](https://github.com/pm0kjp/Philadelphia_Property_Tax/blob/master/data/worstOffenders.csv) and [in a Google spreadsheet](https://docs.google.com/spreadsheets/d/1m_r_KC2js6xkHndqqGfWzrkgUW4RvgDTti9tUKhVQ0k/), and there's a 
Google Charts visualization on CodePen that uses the Google Spreadsheet copy.
You can see the [code](http://codepen.io/kjoypayton/pen/JbyyJx), or just look at the [visualization](http://codepen.io/kjoypayton/full/JbyyJx/).