---
title: "Philadelphia Property Tax Delinquency: Investor Impact"
author: "Joy Payton"
date: "11/30/2016"
output: html_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
```

## Load Libraries

```{r}
library(dplyr)
library(stringr)
library(jsonlite)
```

## Obtain Tax Data 

We can get property tax delinquency data from Philadelphia's Open Data site:

```{r cars}
csv <- "https://data.phila.gov/api/views/4v5p-qi2q/rows.csv?accessType=DOWNLOAD&bom=true"
taxData <- read.csv(csv, stringsAsFactors = FALSE)
```

## Clean Data

Remove columns that don't interest us:

```{r}
taxDataCleaned <- taxData %>% select(-c(Unit,Lien.Number))
```

Remove rows that won't help us because they list no owner:

```{r}
taxDataCleaned <- taxDataCleaned %>% filter(Owner != "")
```

Clean out the $ found in the money fields and convert to numeric

```{r}
taxDataCleaned$Principal <- as.numeric(gsub('\\$','',taxDataCleaned$Principal))
taxDataCleaned$Interest <- as.numeric(gsub('\\$','',taxDataCleaned$Interest))
taxDataCleaned$Penalty <- as.numeric(gsub('\\$','',taxDataCleaned$Penalty))
taxDataCleaned$Other <- as.numeric(gsub('\\$','',taxDataCleaned$Other))
taxDataCleaned$Total <- as.numeric(gsub('\\$','',taxDataCleaned$Total))
```

Get just 2000-2015 tax amount owed.  2016 represents bills that may not truly
be in delinquency yet, and prior to 2000 is too far back for us to be interested.

```{r}
lastFifteenData <- taxDataCleaned %>% filter(Tax.Period < 2016, Tax.Period >= 2000)
```

## What is the impact of investors?

As a secondary exploration into our data, we'll want to see what the impact of
investors is (we saw some interesting trends of large investors in our ["Worst
100 Tax Delinquents" visualization](http://codepen.io/kjoypayton/full/JbyyJx/)).

Let's limit our investigation to the last year - 2015.  We'll take only those
properties that have overdue tax and do some summarizing.  We consider a 
"deadbeat investor" to be someone with three or more overdue properties.

```{r}
investors <- lastFifteenData %>% 
                    filter(Tax.Period == 2015, Total >0) %>% 
                    group_by(Owner) %>% 
                    summarise(TotalOwed = sum(Total), 
                              NumOverdueProperties = n()) %>%
                    arrange(desc(NumOverdueProperties)) %>% 
                    filter(NumOverdueProperties >=3)
```

We have 1449 potential investors... but we know that some are actually 
governmental agencies.  Let's remove those as we discover them:

```{r}
investors <- investors %>% filter(Owner != "PHILA HOUSING AUTHORITY")
investors <- investors %>% filter(Owner != "CITY OF PHILADELPHIA")
investors <- investors %>% filter(Owner != "ADMIN OF VET AFFAIRS")
investors <- investors %>% filter(Owner != "OFFICE OF THE DISTRICT AT")
investors <- investors %>% filter(Owner != "ADMIN OF VETS AFFAIRS")
investors <- investors %>% filter(Owner != "CITY OF PHILA")
investors <- investors %>% filter(Owner != "ADMIN OF VETERAN AFFAIRS")
investors <- investors %>% filter(Owner != "ADM OF VET AFFAIRS")
```

What's the impact of deadbeat investors?

```{r}
sum(investors$TotalOwed)
sum(investors$NumOverdueProperties)
```

We have $5.6 million due on 6500 properties... a huge impact.  We want to 
concentrate on these owners who are unlikely (due to the number of properties)
to be simple homeowners down on their luck.  Let's dig a bit deeper.  

We wonder where these investor-owned deadbeat properties are distributed across 
the city. Our data doesn't include zip code, but it does have geocoded 
coordinates.  We can do something with that!

Let's look at Census tracts.  These are helpful because they are small enough to
give us some neighborhood-by-neighborhood comparison of how many investor-
owned properties are in delinquency.  Because the Census Bureau also collects a
number of demographic data on the tract level, we can also look to see if there
are any patterns.  Do deadbeat investors take on properties in economically 
strapped neighborhoods? 

First, let's get all of our investment properties, and set up our latitude and 
longitude fields. Once again, let's only worry about 2015.

```{r}
investmentProperties <- taxDataCleaned %>% 
                        filter(Tax.Period == 2015, Total >0) %>% 
                        filter (Owner %in% investors$Owner)

investmentProperties$lat <- as.numeric(str_match(investmentProperties$Coordinates, 
                                               "\\((\\d+.\\d+),")[,2])
investmentProperties$long <- as.numeric(str_match(investmentProperties$Coordinates, 
                                                ",\\s+(.+)\\)")[,2])
```

Some of those properties might have the same latitude and longitude, and don't 
need to be reverse geocoded twice.  Let's get the definitive list of unique 
locations to look up tracts for:

```{r}
toGeocode <- investmentProperties[!duplicated(investmentProperties$Coordinates),]
```


Now let's use an FCC API to get the census tract for this latitude and 
longitude!

```{r}
getTract <- function(latitude, longitude)
{
  tract <- vector(length = length(latitude))
  for (i in 1:length(latitude)) {
    lookup <- paste("http://data.fcc.gov/api/block/find?format=json&latitude=", 
                    latitude[i], "&longitude=", longitude[i],
                    sep="")
    response <- fromJSON(txt=lookup)
    tract[i] <- response$Block$FIPS
  }
  return(tract)
}
```

This next bit takes awhile!  We can do it in batches of 1000 to give a tiny 
bit of throttling...  We'll save it for use later also!

```{r}
toGeocode$tract<-NA
toGeocode$tract[1:1000]<-getTract(toGeocode$lat[1:1000], toGeocode$long[1:1000])
toGeocode$tract[1001:2000]<-getTract(toGeocode$lat[1001:2000], toGeocode$long[1001:2000])
toGeocode$tract[2001:3000]<-getTract(toGeocode$lat[2001:3000], toGeocode$long[2001:3000])
toGeocode$tract[3001:4000]<-getTract(toGeocode$lat[3001:4000], toGeocode$long[3001:4000])
toGeocode$tract[4001:5000]<-getTract(toGeocode$lat[4001:5000], toGeocode$long[4001:5000])
toGeocode$tract[5001:6000]<-getTract(toGeocode$lat[5001:6000], toGeocode$long[5001:6000])
toGeocode$tract[6001:length(toGeocode$tract)]<-getTract(toGeocode$lat[6001:length(toGeocode$tract)], toGeocode$long[6001:length(toGeocode$tract)])
```

The code that's passed in includes precision at the sub-tract level we don't 
care about.  Let's fix that, and save the coordinate / tract table so that we
don't have to do that crazy long API call series again!

```{r}
write.csv(toGeocode[,c('lat','long','tract')],"censusTractsByLatLong.csv",row.names = FALSE)
```

If we're re-running this script, we may want to skip the API calls of every property and do the following instead, to just geocode the ones that need it:

```{r}
toGeocode <- investmentProperties[!duplicated(investmentProperties$Coordinates),]
tractsIKnow<-read.csv("https://raw.githubusercontent.com/pm0kjp/Philadelphia_Property_Tax/master/data/censusTractsByLatLong.csv")
toGeocode <- merge(x=toGeocode, y=tractsIKnow, 
                   by=c('lat','long'), 
                   all.x=TRUE, sort=FALSE)
tractsIDontKnow <- toGeocode[is.na(toGeocode$tract),]
toGeocode <- toGeocode[!is.na(toGeocode$tract),]
tractsIDontKnow <- tractsIDontKnow %>% filter(!is.na(lat),!is.na(long))
tractsIDontKnow$tract<-getTract(tractsIDontKnow$lat, tractsIDontKnow$long)
toGeocode <- bind_rows(toGeocode, tractsIDontKnow)
# Now we have an even richer dataset, and I can replace my GitHub file
# with a new CSV!
write.csv(toGeocode[,c('lat','long','tract')],"censusTractsByLatLong.csv",row.names = FALSE)
```

Either way, I need to trim my tracts, because there's too much precision.
I only want the first 11 numbers:

```{r}
toGeocode$trimmedTract <- strtrim(toGeocode$tract,11)
```

Alright!  Now we can make the connection to all of the properties across all 
the years of data we have:

```{r}
investmentProperties <- merge(investmentProperties, 
                              toGeocode[,c('lat','long','trimmedTract')], 
                              by=c("lat","long"))
```

We now have census tracts for all our investment properties. For our D3 
visualization, we want to show the total 2015 amount of deadbeat investor unpaid 
tax per census tract.

```{r}
investmentPropertiesByTract2015 <- investmentProperties %>%
  filter(Tax.Period == 2015, Total >=0) %>%
  group_by(trimmedTract) %>%
  summarise(totalOwedByTract = sum(Total))
```

Now we will take the Census Tract GeoJSON file stashed in GitHub, and enrich its
"properties" section with our summarized data about amount owed per tract. We'll
make sure that if there's nothing showing as owed on that tract, we put a zero
in.

```{r}
tractData <- fromJSON("https://raw.githubusercontent.com/pm0kjp/Philadelphia_Property_Tax/master/data/Census_Tracts_2010.geojson")
enrichedProperties <- merge(x=tractData$features$properties, y=investmentPropertiesByTract2015, by.x="GEOID10", by.y="trimmedTract", all.x=TRUE)
enrichedProperties$totalOwedByTract[is.na(enrichedProperties$totalOwedByTract)] <- 0
```

We can also enrich our GeoJSON with economic data about the tract generally, 
taken from the Census Bureau itself:

```{r}
economicData <- read.csv("https://raw.githubusercontent.com/pm0kjp/Philadelphia_Property_Tax/master/data/ACS_10_5YR_DP03.csv", stringsAsFactors = FALSE)

economicDataCleaned <- economicData %>% select(GEO.id2, HC03_VC166, HC01_VC85)

names(economicDataCleaned) <- c('GEOID10', 'PctPoverty','MedianHouseholdIncome')

enrichedProperties <- merge(x=enrichedProperties, y=economicDataCleaned, by="GEOID10", all.x=TRUE)
```

We'll replace the original "properties" in our GeoJSON with this enriched
"properties", and make a new JSON file:

```{r}
tractData$features$properties <- enrichedProperties

myGeoJSON <- toJSON(tractData)

write(myGeoJSON, file="enrichedPhilly.geojson")
```